Correctly Optimizing Abstract Abstract Machines

### 2013-03-29 paper/paper.tex
More text fix ups.

%\title{Provably Correct, Low-Cost Optimizations for Abstract Abstract Machines}
\title{Systematically Optimizing Abstract Abstract Machines}

@@

The AAM approach, in essence, does the following: it takes a
machine-based view of computation and turns it into a \emph{finitary
  approximation} by bounding the size of the store.  With a limited
address space, the store must map addresses to \emph{sets} of values.
Store updates are interpreted as joins, and store dereferences are
interpreted by non-deterministic choice of an element from a set.  The
result of analyzing a program is a finite directed graph where nodes
in the graph are machine states and edges denote machine
transitions between states.

@@

%\newpage
\section{An analyzer for ISWIM}
\label{sec:aam}

@@

The semantics are defined in terms of a machine model.  The machine
components are defined in figure~\ref{fig:domains} and
figure~\ref{fig:aam} defines the transition relation.  The evaluation
of a program is defined as its traces given by the machine transition relation.  The
machine is a very slight variation on a standard abstract machine for
ISWIM in ``eval, continue, apply'' form~\cite{dvanhorn:Danvy:DSc}.  It
can be systematically derived from a definitional interpreter through
a continuation-passing style transformation and defunctionalization,
or from a structural operational semantics using the refocusing
construction of Danvy and
Nielsen~\cite{dvanhorn:Danvy-Nielsen:RS-04-26}.

Compared with the standard machine semantics, this definition is
different in the following ways, which make it abstractable as a
program analyzer:
\begin{itemize}
\item the store maps addresses to \emph{sets} of values, not
  single values,
\item continuations are heap-allocated, not stack-allocated,
\item there are ``timestamps'' ($\mcntr \in \Counter$) and syntax
  labels ($\mlab$) threaded through the computation, and
\item the machine is implicitly parameterized by the functions
  $\alloc$, $\allockont$, $\tick$, $\interpdelta$, and spaces
  $\Addr$, $\Counter$ (and initial $\mcntr_0 \in \Counter$).
\end{itemize}

More generally, we can make stores map to a domain that forms a galois
connection with sets of values, so that $\interpdelta$ can give more
elaborate abstractions of base values (e.g., with the interval
abstraction, $\{0,[0,5]\} \equiv \{[0,5]\}$). We stick with sets for
simpler exposition.

@@

\paragraph{Concrete interpretation} To characterize concrete interpretation, set the implicit
parameters of the relation given in figure~\ref{fig:aam} as follows:
\begin{align*}
\alloc(\mstate) &= \maddr \mbox{ where } \maddr \notin \mstate.\msto \\
\allockont^\mcntr_\mlab(\msto,\mkont) \mbox{ where } \maddr \notin \msto
\end{align*}
where $\sqcup$ on stores is a point-wise lifting of $\sqcup$: $\msto
\sqcup \msto' = \lambda \maddr. \msto(\maddr) \sqcup
\msto'(\maddr)$. These functions appear to ignore $\mlab$ and
$\mcntr$, but they can be used to determinize the choice of fresh
addresses.  The resulting relation is non-deterministic in its choice
of addresses, however it must always choose a fresh address when
allocating a continuation or variable binding.  If we consider machine
states equivalent up to consistent renaming and fix an allocation
scheme, this relation defines a deterministic machine (the relation is
really a function).

@@

\paragraph{Abstract interpretation} To characterize abstract
interpretation, set the implicit parameters just as above, but drop
the $\maddr \not\in \msto$ condition. $\interpdelta$ takes some care
to not make the analysis run forever; a simple instantiation is a flat
abstraction where arithmetic operations return an abstract top element
$\sNum$, and $\szerohuh$ returns both $\strue$ and $\sfalse$ on
$\sNum$.  This family of interpreters is also non-deterministic in
choices of addresses, but it is free to choose addresses that are
already in use.  Consequently, the machines may be non-deterministic
when multiple values reside in a store location.

It is important to recognize from this definition that \emph{any}
allocation strategy is an abstract
interpretation~\cite{dvanhorn:Might2009Posteriori}.  In particular,
concrete interpretation is a kind of abstract interpretation.  So is
an interpretation that allocates a single cell into which all bindings
and continuations are stored.  On the one hand is an abstract
interpretation that is non-computable and gives only the ground truth
of a programs behavior; on the other is an abstract interpretation
that is easy to compute but gives little information.  Useful program
analyses lay somewhere in between and can be characterized by their
choice of address representation and allocation strategy.  Uniform
\(k\)-CFA~\cite{dvanhorn:nielson-nielson-popl97} is one such analysis.

@@

\section{Reduction semantics to baseline analyzer}
\label{sec:baseline}

The uniform $k$-CFA allocation strategy would make $\traces$ in figure
\ref{fig:aam} a computable abstraction of possible executions, but not an
efficient one. This is not the strategy that AAM, nor we, recommend. Through
this section, we will explain a succession of approximations
to reach the baseline analysis.  We'll compare performance at each stage to
identify the criticality of each choice.
%
We ground this journey by first formulating the analysis in terms of a classic
fixed-point computation.

@@

In the next subsection, we'll fix this with a widening and reach polynomial
(albeit of a high degree) complexity.
%
This widening effectively lifts the store out of individual states to create
a single, global shared store for all.

@@

A common technique to accelerate convergence in flow analyses is to share a
common, global store.
%
To retain soundness, this store grows monotonically.
%
Formally, we can cast this optimization as a second abstraction or as the
application of a widening operator during the fixed-point iteration.
%
In the ISWIM language, such a widening makes 0-CFA quartic in the size of the
program.
%
Thus, in one step, complexity drops from intractable exponentiality to a merely
daunting polynomial.

@@

To describe this process, AAM defined a transformation of the reduction relation so that it operates on
a pair of a set of contexts ($C$) and a store ($\sigma$).
%
A context includes all non-store components, \emph{e.g.}, the expression, the environment and the stack.
%
The transformed relation, $\widehat{\machstep}$, is
%
\begin{align*}
(C, \msto) &\mathrel{\widehat{\machstep}} (C', \msto') \\
\mbox{where } C' &= \{c' \mid \wn(c, \msto) \mathrel{\machstep} \wn(c', \msto^c), c \in C\} \\
              \msto' &= \bigsqcup\; \{\msto^c \mid \wn(c,\msto)\mathrel{\machstep} \wn(c', \msto^c), c \in C\} \\
\wn &: \widehat{\State} \times \Store \to \State \\
\wn(\ev{\mexp, \menv, \mkont}, \msto) &= \ev{\mexp, \menv, \msto, \mkont} \\
\wn(\co{\mval, \mkont}, \msto) &= \co{\mval, \mkont, \msto} \\
\wn(\ap{\mvalx{u}, \mval, \mkont}, \msto) &= \ap{\mvalx{u}, \mval, \msto, \mkont} \\
\wn(\ans{\mval}, \msto) &= \ans{\msto, \mval}
\end{align*}
%
In effect, the new store is computed as the least upper bound of all
stores after stepping. For clarity, we will exhibit non-widened
semantics unless a technique specifically requires it. The final step
to the baseline takes the complexity from quartic to cubic in the
monovariant case.

@@

The final approximation we make to get to our point of departure is to
store-allocate all values that appear, so that any non-machine state
that contains a value instead contains an address to a value.  The AAM
approach stops at the previous optimization.  However, the {\tt fn}
continuation stores a value, and this makes the space of continuations
quadratic rather than linear in the size of the program, for a
monovariant analysis like 0-CFA.  Having the space of continuations
grow linearly with the size of the program will drop the overall
complexity to cubic (as expected).

To achieve this linearity for continuations, we allocate an address
for the value position when we create the continuation.  This address
and the tail address are both determined by the label of the
application point, so the space becomes linear and the overall
complexity drops to cubic.  This is a critical abstraction in
languages with $n$-ary functions, since otherwise the continuation
space grows super-exponentially. We extend the semantics to
additionally allocate an address for the function value when creating
the $\kfn{}$ continuation. The continuation has to store this address
to remember where to get values from, though sometimes your allocation
strategy may allow you to elide that representation.

The new evaluation rules follow (where $\mcntr' = \tick(\mstate)$).
% In theory, this aggressive distinction between continuations might buy
% additional precision, but in practice, it does not.

>>>>

%\title{Provably Correct, Low-Cost Optimizations for Abstract Abstract Machines}
\title{Correctly Optimizing Abstract Abstract Machines}

@@

The AAM approach, in essence, does the following: it takes a
machine-based view of computation and turns it into a \emph{finitary
  approximation} by bounding the size of the store.  With a limited
address space, the store must map addresses to \emph{sets} of values.
Store updates are interpreted as joins, and store dereferences are
interpreted by non-deterministic choice of an element from a set.  The
result of analyzing a program is a finite directed graph where nodes
in the graph are (abstract) machine states and edges denote machine
transitions between states.

@@

%\newpage
\section{Abstract interpretation of ISWIM}
\label{sec:aam}

@@

The semantics are defined in terms of a machine model.  The machine
components are defined in figure~\ref{fig:domains};
figure~\ref{fig:aam} defines the transition relation.  The evaluation
of a program is defined as its set of traces that arise from iterating the
machine transition relation.  The
machine is a very slight variation on a standard abstract machine for
ISWIM in ``eval, continue, apply'' form~\cite{dvanhorn:Danvy:DSc}.  It
can be systematically derived from a definitional interpreter through
a continuation-passing style transformation and defunctionalization,
or from a structural operational semantics using the refocusing
construction of Danvy and
Nielsen~\cite{dvanhorn:Danvy-Nielsen:RS-04-26}.

Compared with the standard machine semantics, this definition is
different in the following ways, which make it abstractable as a
program analyzer:
\begin{itemize}
\item the store maps addresses to \emph{sets} of values,\footnote{
More generally, we can have stores map to any domain that forms
 a Galois connection with values, enabling $\interpdelta$ to
produce elaborate abstractions of base values (e.g., interval or octagon
abstractions). We use sets of values for a simpler exposition.
} not
  single values,

\item continuations are heap-allocated, not stack-allocated,
\item there are ``timestamps'' ($\mcntr \in \Counter$) and syntax
  labels ($\mlab$) threaded through the computation, and
\item the machine is implicitly parameterized by the functions
  $\alloc$, $\allockont$, $\tick$, $\interpdelta$, and spaces
  $\Addr$, $\Counter$ (and initial $\mcntr_0 \in \Counter$).
\end{itemize}

@@

\paragraph{Concrete interpretation} To characterize concrete interpretation, set the implicit
parameters of the relation given in figure~\ref{fig:aam} as follows:
\begin{align*}
\alloc(\mstate) &= \maddr \mbox{ where } \maddr \notin \mstate.\msto \\
\allockont^\mcntr_\mlab(\msto,\mkont) &= \maddr \mbox{ where } \maddr \notin \msto
\end{align*}
where $\sqcup$ on stores is a point-wise lifting of $\sqcup$, \ie,
$\msto \sqcup \msto' = \lambda \maddr. \msto(\maddr) \sqcup
\msto'(\maddr)$ and $\mstate.\msto$ is notation for projecting out the
store component of a state. These functions appear to ignore $\mlab$
and $\mcntr$, but they can be used to determinize the choice of fresh
addresses.  The resulting relation is non-deterministic in its choice
of addresses, however it must always choose a fresh address when
allocating a continuation or variable binding.  If we consider machine
states equivalent up to consistent renaming and fix an allocation
scheme, this relation defines a deterministic machine (the relation is
really a function).

@@

\paragraph{Abstract interpretation} To characterize abstract
interpretation, set the implicit parameters just as above, but drop
the $\maddr \not\in \msto$ condition. The $\interpdelta$ relation takes some care
to not make the analysis run forever; a simple instantiation is a flat
abstraction where arithmetic operations return an abstract top element
$\sNum$, and $\szerohuh$ returns both $\strue$ and $\sfalse$ on
$\sNum$.  This family of interpreters is also non-deterministic in
choices of addresses, but it is free to choose addresses that are
already in use.  Consequently, the machines may be non-deterministic
when multiple values reside in a store location.

It is important to recognize from this definition that \emph{any}
allocation strategy is a sound abstract
interpretation~\cite{dvanhorn:Might2009Posteriori}.  In particular,
concrete interpretation is a kind of abstract interpretation.  So is
an interpretation that allocates a single cell into which all bindings
and continuations are stored.  The former is an abstract
interpretation that is non-computable and gives only the ground truth
of a programs behavior; the latter is an abstract interpretation
that is easy to compute but gives little information.  Useful program
analyses lay somewhere in between and can be characterized by their
choice of address representation and allocation strategy.  Uniform
\(k\)-CFA~\cite{dvanhorn:nielson-nielson-popl97}, presented next, is one such analysis.

@@

\section{From machine semantics to baseline analyzer}
\label{sec:baseline}

The uniform $k$-CFA allocation strategy would make $\traces$ in figure
\ref{fig:aam} a computable abstraction of possible executions, but one
that is too inefficient to run, even on small examples.  Through this
section, we explain a succession of approximations to reach a more
appropriate baseline analysis.
%
We ground this path by first formulating the analysis in terms of a
classic fixed-point computation.

@@

In the next subsection, we fix this with store widening to reach polynomial
(albeit of high degree) complexity.
%
This widening effectively lifts the store out of individual states to create
a single, global shared store for all.

@@

A common technique to accelerate convergence in flow analyses is to share a
common, global store.
%
To retain soundness, this store grows monotonically.
%
Formally, we can cast this optimization as a second abstraction or as the
application of a widening operator during the fixed-point iteration.
%
In the ISWIM language, such a widening makes 0-CFA quartic in the size of the
program.
%
Thus, complexity drops from intractable exponentiality to a merely
daunting polynomial.

@@

To describe this process, AAM defined a transformation of the reduction relation so that it operates on
a pair of a set of contexts ($C$) and a store ($\sigma$).
%
A context includes all non-store components, \emph{e.g.}, the expression, the environment and the stack.
%
The transformed relation, $\widehat{\machstep}$, is
%
\begin{align*}
(C, \msto) &\mathrel{\widehat{\machstep}} (C', \msto'), \\
\mbox{where } C' &= \{c' \mid \wn(c, \msto) \mathrel{\machstep} \wn(c', \msto^c), c \in C\} \\
              \msto' &= \bigsqcup\; \{\msto^c \mid \wn(c,\msto)\mathrel{\machstep} \wn(c', \msto^c), c \in C\} \\
\wn &: \widehat{\State} \times \Store \to \State \\
\wn(\ev{\mexp, \menv, \mkont}, \msto) &= \ev{\mexp, \menv, \msto, \mkont} \\
\wn(\co{\mval, \mkont}, \msto) &= \co{\mval, \mkont, \msto} \\
\wn(\ap{\mvalx{u}, \mval, \mkont}, \msto) &= \ap{\mvalx{u}, \mval, \msto, \mkont} \\
\wn(\ans{\mval}, \msto) &= \ans{\msto, \mval}
\end{align*}
%
In effect, the new store is computed as the least upper bound of all
stores after stepping. For clarity, we will exhibit non-widened
semantics unless a technique specifically requires it. The final step
to the baseline takes the complexity from quartic to cubic in the
monovariant case.

@@

The final approximation we make to get to our baseline is to
store-allocate all values that appear, so that any non-machine state
that contains a value instead contains an address to a value.  The AAM
approach stops at the previous optimization.  However, the {\tt fn}
continuation stores a value, and this makes the space of continuations
quadratic rather than linear in the size of the program, for a
monovariant analysis like 0-CFA.  Having the space of continuations
grow linearly with the size of the program will drop the overall
complexity to cubic (as expected).

To achieve this linearity for continuations, we allocate an address
for the value position when we create the continuation.  This address
and the tail address are both determined by the label of the
application point, so the space becomes linear and the overall
complexity drops to cubic.  This is a critical abstraction in
languages with $n$-ary functions, since otherwise the continuation
space grows super-exponentially. We extend the semantics to
additionally allocate an address for the function value when creating
the $\kfn{}$ continuation. The continuation has to contain this address
to remember where to retreive values from in the store.

The new evaluation rules follow, where $\mcntr' = \tick(\mstate)$:
% In theory, this aggressive distinction between continuations might buy
% additional precision, but in practice, it does not.
